A Collection of Thoughts on the Keccak/SHA-3 Construction

### 2018-05-01 s2018-math490/thesis/thesis.tex
Finished thesis
Signed-off-by: Alexander Scheel <alexander.m.scheel@gmail.com>

\begin{abstract}
    % TODO
    TODO
\end{abstract}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Bergman and utilized
the resulting
\href{https://github.com/cipherboy/hash\_framework}{hash\_framework} project.
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.

@@

@@

    Further, a cryptographic hash function must be resistant finding a
preimage: for a given output value $v$, finding an input $x$ such that
$f(x) = v$ should be roughly on the order of $2^{b}$.

@@

    FIPS 202 \cite{NIST202}.

@@

\subsection{Introduction on the Structure of SHA-3} \label{sec:i:structure}

    SHA-3 consists of two parts: a core permutation function, KECCAK-$f$, and
a domain extender, the KECCAK sponge function. As discussed by the Keccak
authors, additional security is given by choosing the permutation functions to
be bijective, though they need not be (the domain need only be $S_{w}$).

@@

\section{Marginal and Differential Properties of the Five Round Functions} \label{sec:differentials}

This section studies three approaches to extracting information about the five
permutations functions using SAT. None of these approaches can directly lead
to a collision in SHA-3, however, by combining them, possible differential
trails can be approximated. Sadly for $w \geq 4$, several of these attacks take
too long to run (even for individual permutation functions) and thus do not
emit a valid attack for full Keccak.

@@

\subsection{Marginal Properties} \label{sec:d:margin}

@@

\subsection{Differential Properties} \label{sec:d:diff}

@@

\subsection{Input Margin Impact on Differential Properties} \label{sec:d:input}

@@

\subsection{Output Margin Impact on Differential Properties} \label{sec:d:output}

@@

Future work includes scaling the partial fixed point attacks via intermediate
models, evaluating weaknesses in alternate constructions, and improving
the run time of marginal and differential analysis. However, without
significant breakthroughs in logical cryptanalysis, most of these attacks are
unlikely. Further, it is not immediately obvious what effect low permutation
order has on the possibility of collisions; introducing permutation-based
cryptanalysis could result in new techniques applicable to SHA-3.

>>>>

\begin{abstract}
    % TODO
\end{abstract}

@@

\section{A Series of Introductions} \label{sec:intro}
    This paper presents the work of the author towards the completion of the
requirements for the MATH 490 Independent Study course under Dr. Clifford
Bergman at Iowa State University of Science and Technology. This work was an
extension of the author's Honors Project under Dr. Eric Bergman and utilized
the resulting
\href{https://github.com/cipherboy/hash\_framework}{hash\_framework} project.
Additional artifacts related to this project can be seen in the
\href{https://github.com/cipherboy/keccak-attacks}{keccak-attacks} repository.
A permanent location for this document is in the
\href{https://github.com/cipherboy/papers}{papers} repository.

@@

    The remaining sections outside of the introductions discuss 1) various
mathematical properties of the core round functions, 2) exhaustive collision
searches on small instances, 3) correlation matrices, and 4) fixed point
attacks.

@@

    Further, a cryptographic hash function must be resistant finding a
preimage: for a given output value $v$, finding an input $x$ such that
$f(x) = v$ should be roughly on the order of $2^{b}$. Given such an attack,
it is possible to find a collision: choose a random input, compute its hash,
then use the preimage attack to find a different input for that hash; these two
inputs form a collision pair. Thus a preimage attack is stronger than a
collision attack.

    However, if the attacker is allowed to choose only one message (that is,
the challenger chooses the other), this forms a second preimage attack. More
formally, for a given fixed message $x$, the attacker seeks a $y \neq x$ such
that $h(x) = h(y)$. While a collision attack is only valid for some such
messages, $x$, a preimage attack has to be valid for all such $x$; thus it is
a stronger attack than collision attacks, but weaker than a preimage attack
as it presupposes an existing message.

    This independent study focused on a single hash function, Keccak, which
was chosen by NIST to be the next standardized algorithm, SHA-3 \cite{NIST202}.
This hash function differs from previously standardized algorithms in two major
ways: one, it is not based on the Merkel-Damg{\aa}rd construct and is rather
based on the Sponge construction, and two, the internal round function is not a
compression function and is instead a permutation. Outside of a few theorems
done by hand and verifying the lack of low-order correlations, we focused on
using automated tools to learn more about the structure of SHA-3.

    In the late 1990s, F. Massacci started analyzing cryptographic
constructions using Boolean Satisfability \cite{CryptoSAT}; however, applying
this to hash functions is a largely understudied field. E. Homsirikamol tried
brute forcing collisions in SHA-3 to limited results \cite{Homsirikamol2012},
and P. Morawiecki performed similar work for preimages \cite{cryptoeprint:2010:285}.
However, neither contributed useful, general cryptanalysis techniques to scale
attacks against reduced rounds (and reduced sizes) to larger numbers of rounds.
Our work begins exploring new techniques and several consequences in the
following sections.

@@

@@

    NIST standardized Keccak as SHA-3 in FIPS 202 \cite{NIST202}; Keccak was
chosen as the winner of the corresponding hash function design competition
started in 2009. Designed by the Keccak Team (G. Bertoni, J. Daemen, M.
Peeters, G. Van Assche, and R. Van Keer \cite{KeccakTeam}), Keccak features a
novel construction.

    Keccak is built around the concept of a state cube: a 5 by 5 by $w$ cube
of bits (where $w$ is a power of 2). By increasing $w$, a higher security
margin can be given, at the expense of increasing the computational intensity.
However, because internal round operations on this state cube are bijective,
it can be implemented efficiently in hardware with minimal gate and propagation
delays. Typically this state is called $A$ when represented as a cube, and $s$
when represented as a (flat) bit string. There are five major functions which
compose to form a single round: $\theta$, $\rho$, $\pi$, $\chi$, and $\iota$;
each maps bijectively over the entire state space (i.e., elements of
$2^{25w}$).

    $\theta$ is an eleven way XOR and is thus a linear map. For all
$0 \leq x \leq 4$ and $0 \leq z < w$, define:
\begin{align*}
    C[x, z] & := \oplus_{0 \leq y \leq 4} A[x, y, z] \\
    D[x, z] & := C[(x-1) \text{ mod } 5, z] \oplus C[(x+1) \text{ mod } 5, (z - 1) \text{ mod } w] \\
\end{align*}
Then:
\begin{align*}
    A'[x, y, z] & := A[x, y, z] \oplus D[x, z]
\end{align*}

    $\rho$ is a pure permutation function, permuting the locations of bits in
a lane (the $z$ component) of the state cube. For all
$0 \leq z < w: A'[0, 0, z] := A[0, 0, z]$. Then let $x, y = (0, 0)$, and for
$0 \leq t \leq 23$:
\begin{align*}
    \forall 0 \leq z < w: A'[x, y, z] & := A[x, y, (z - \frac{(t+1)(t+2)}{2}) \text{ mod } w] \\
    let x, y & = (y, (2x + 3y) \text{ mod } 5)
\end{align*}

    $\pi$ is another pure permutation function, permuting locations along the
face of the cube. For all $0 \leq x \leq 4, 0 \leq y \leq 4, 0 \leq z < w$:
\begin{align*}
    A'[x, y, z] := A[(x + 3y) \text{ mod } 5, x, z].
\end{align*}

    $\chi$ is the only non-linear function and is of degree one (with inverse
of degree three). For all $0 \leq x \leq 4, 0 \leq y \leq 4, 0 \leq z < w$:
\begin{align*}
    A'[x, y, z] := A[x, y, z] \oplus (\lnot A[(x + 1) \text{ mod } 5, y, z] \land A[(x + 2) \text{ mod } 5, y, z]).
\end{align*}

    $\iota$ is a fixed-value XOR useful for preventing trivial fixed points,
and is the only function dependent on the current round number. Refer to
page 16 of FIPS 202 (\cite{NIST202}) for its specification; it is not discussed
extensively in this paper.

    The sponge construction can be described as follows. Fix $w$ as a power of
two. To have a security margin of $m$ bits, the output needs to have $m$ bits;
however, if the attacker has complete control of all $25w$ bits of state,
they would be able to choose the output easily because Keccak is bijective.
Thus, restrict the input to $25w - 2m$ bits and set the remaining $2m$ bits to
zero. To find a preimage for a $m$-bit output, the attacker would have to find
a $25w - m$ bit suffix such that the preimage of that suffix has $2m$ zeros.
Similarly, for a collision attack, an attacker has to find two $25w-2m$
messages such that their hashes have the same prefix of length $m$. Under an
ideal round permutation, clearly these are non-trivial and hard to find in
general.

    For a multiple block message the procedure is thus as follows: call the
composition of 24 rounds to be \texttt{KECCAK-$f$}. Start with $s=0^{25w}$.
Splitting the message into chunks of $25w-2m$ (and padding according to
the scheme defined in FIPS 202 \cite{NIST202}), for each block $b$, extend
$b$ to length $25w$ with zeros, then compute $s = f(s \oplus b)$. From $s$,
take the top $m$ bits as the output of the hash function.

\subsection{Introduction to the Remaining Sections}

%\subsection{Introduction on the Structure of SHA-3} \label{sec:i:structure}

%    SHA-3 consists of two parts: a core permutation function, KECCAK-$f$, and
%a domain extender, the KECCAK sponge function. As discussed by the Keccak
%uthors, additional security is given by choosing the permutation functions to
%be bijective, though they need not be (the domain need only be $S_{w}$).

@@

%\section{Marginal and Differential Properties of the Five Round Functions} \label{sec:differentials}

%This section studies three approaches to extracting information about the five
%permutations functions using SAT. None of these approaches can directly lead
%to a collision in SHA-3, however, by combining them, possible differential
%trails can be approximated. Sadly for $w \geq 4$, several of these attacks take
%too long to run (even for individual permutation functions) and thus do not
%emit a valid attack for full Keccak.

@@

%\subsection{Marginal Properties} \label{sec:d:margin}

@@

%\subsection{Differential Properties} \label{sec:d:diff}

@@

%\subsection{Input Margin Impact on Differential Properties} \label{sec:d:input}

@@

%\subsection{Output Margin Impact on Differential Properties} \label{sec:d:output}

@@

Future work includes scaling the partial fixed point attacks via intermediate
models, evaluating weaknesses in alternate constructions, and improving
the run time of marginal and differential analysis. However, without
significant breakthroughs in logical cryptanalysis, most of these attacks are
unlikely. Further, it is not immediately obvious what effect low permutation
order has on the possibility of collisions; introducing permutation-based
cryptanalysis could result in new techniques applicable to SHA-3. Additional
work could include automatic lemmaizing of found constraints and including
them in future searches.
